---
title: "ODE and Sobel"
author: "Heather Childers & Sofia Ingersoll"
date: "2024-05-17"
output: html_document
---

# Using Sobel with an ODE

## The Model

Consider the following model of forest growth (where forest size in measured in units of carbon (C))
dC/dt = r ∗ C for forests where C is below a threshold canopy closure
dC/dt = g ∗ (1 − C/K) for forests where carbon is at or above the threshold canopy closure
and K is a carrying capacity in units of carbon
The size of the forest (C), Canopy closure threshold and carrying capacity are all in units of carbon You
could think of the canopy closure threshold as the size of the forest at which growth rates change from
exponential to linear You can think of r, as early exponential growth rate and g as the linear growth rate
once canopy closure has been reached

## Homework Task

Start by loading in all the necessary libraries/packages

```{r, message = FALSE}
#Data analysis
library(tidyverse)
library(purrr)
library(deSolve) #ODE Solver
library(sensitivity) #Use to create Sobel analysis
```

1.  Implement this model as a function in R

Great! Let's pull from our R script that we have already made

```{r}
# Source the function from our R script
source("canopy_growth.R")
```

2.  Run the model for 300 years with the specified initial values(Using the ode solver)

```{r}
#let's set some initial conditions
initial_carbon = 10
years = seq_along(1:300)
parameters <- c(r = 0.01, thresh = 50, g = 2, K = 250)

# run the solver
Canopy = ode(y=initial_carbon, times=years, func=canopy_growth, parms = parameters) %>% 
  as.data.frame()  
colnames(Canopy)=c("time","Carbon")

# peek at the df
#head(Canopy, 3)
```

3.  Graph the results from the above calculations

```{r}
Canopy %>% 
  ggplot(aes(x = time, y = Carbon))+
  geom_point(color = 'forestgreen', size = 2)+
  labs(title = "Forest Size (kgC)",
       x = "Time (years)",
       y = "Carbon (kg)")
```

4.  Now lets run a sensitivity analysis for our parameters

```{r}
#------------------------------------------------
#----    Create our parameter sets           ----
#------------------------------------------------
# want to learn about sensitivity to growth rates (r & g) and carrying capacity (K)
# set the number of parameters
np=2000
K = rnorm(mean=250, sd=25, n=np)
r = rnorm(mean=0.01, sd=0.001, n=np)
g = rnorm(mean=2, sd=0.2, n=np)
X1 = cbind.data.frame(r=r, K=K, g=g)

np=2000
K = rnorm(mean=250, sd=25, n=np)
r = rnorm(mean=0.01, sd=0.001, n=np)
g = rnorm(mean=2, sd=0.2, n=np)
X2 = cbind.data.frame(r=r, K=K, g=g)

# fix any negative values and they are not meaningful
X1 = X1 %>% map_df(pmax, 0.0)
X2 = X2 %>% map_df(pmax, 0.0)

#------------------------------------------------
#----          Create our sobel object       ----
#------------------------------------------------
# create our sobel object and get sets of parameters for running the model
sens_C = sobolSalt(model = NULL,X1,X2, nboot = 300)

# our parameter sets are
head(sens_C$X)
# lets add names 
colnames(sens_C$X) = c("r","K",'g')

#------------------------------------------------
#----    Run the ODE Solver with Sobel se    ----
#------------------------------------------------
# Create a function to compute the maximum forest size
compute_metrics = function(result){
  maxsize = max(result$C)
  return(list(maxsize=maxsize))}

# make a wrapper function to run the ODE solver with the full Sobel set
p_wrapper = function(r,K,g,thresh, C, time, odefunc, metricfunc) {
  parms = list(r=r, thresh=thresh, K=K, g=g)
  result = ode(y=C, times=time, func=odefunc, parms=parms) 
  colnames(result)=c("time","C")
  # get metrics
  metrics=metricfunc(as.data.frame(result))
  return(metrics)
}

# now use pmap to run the wrapper function
allresults = as.data.frame(sens_C$X) %>% pmap(p_wrapper, thresh = 50, C= initial_carbon, time = years, odefunc=canopy_growth, metricfunc=compute_metrics)

# extract out results from pmap into a data frame
allres = allresults %>% map_dfr(`[`,c("maxsize"))
```

5.  Graph the results and record the S & T Sobel indices

From what we are seeing based on the Sobel indices, the linear growth rate that occurs after the threshold is reached is not as important in determining the max forest size as the exponential growth rate and the carrying capacity. When examining the sensitivity analysis indices in column 1, it appears that the exponential growth rate (r) and the carrying capacity (K) have a more significant impact on determining the maximum forest size compared to the linear growth rate (g) after the threshold is reached. This is important as the climate continues to change because I would anticipate that climate change will cause the carrying capacity to decrease. As temperatures rise and weather patterns become more erratic, the carrying capacity of forests is expected to decrease. As we get hotter summers and wetter winters, the number of trees a forest can support will be limited to how many can survive through the hot summer. This logic also applies to the exponential growth rate (r), as it is likely to be highly variable in response to changing seasonal conditions. As we get hotter summers and wetter winters, the exponential growth rate will vary drastically based on the season. This will seriously impact maximum forest size as the exponential growth rate is the most influential parameter that we tested. In conclusion, the most impactful and influential parameters in this model are (r) and (K), these two parameters will be the key drivers in for understanding and predicting changes in forest size under changing environmental conditions. 

```{r}
# create boxplot of max forest size
tmp = allres %>% pivot_longer(cols=everything(),names_to="metric", values_to="value") 
ggplot(tmp, aes(metric, value, col=metric))+
  geom_boxplot(color = 'forestgreen')+
  labs(title = "Distribution of Maximum Forest Size (kgC)",
       x = "",
       y = "Maximum Forest Size (kgC)")
```

```{r}
# Record the Sobel indeces
sens_C_maxsize = sensitivity::tell(sens_C,allres$maxsize)

# first-order indices (main effect without co-variance)
rownames(sens_C_maxsize$S)=c("r","K","g")
sens_C_maxsize$S

# total sensitivity index -note that this partitions the output variance 
rownames(sens_C_maxsize$T)=c("r","K", "g")
sens_C_maxsize$T

```